---
title: "Hands-on Exercise 1: Geospatial Data Wrangling  with R"
author: "Hulwana"
editor: visual
---

## 1 Overview

In this hands-on exercise, I learn how to import and wrangle geospatial data using appropriate R packages.

## 1.1 Getting Started

The code chunk below install and load the following packages into R environment.

-   [**sf**](https://r-spatial.github.io/sf/) : for importing, managing and processing geospatial data

-   [**tidyverse**](https://www.tidyverse.org/) : for data preparation such as importing, wrangling and visualizing data

```{r}
pacman::p_load(sf, tidyverse)
```

## 1.2 Importing Geospatial Data

Unlike general data saved in excel or csv, geospatial data has to be read by the *st_read()* function of the **sf** package to capture the polygon feature. An example of importing geospatial data is as shown below.

### 1.2.1 Importing polygon feature data in shapefile format

```{r}
mpsz <- st_read(dsn = "data/geospatial", layer = "MP14_SUBZONE_WEB_PL")
```

The above message indicates that the geospatial data consists of 323 multipolygon features and 15 fields.

### 1.2.2 Importing polyline feature data in shapefile form

The code chunk below uses *st_read()* function of **sf** package to import `CyclingPath` shapefile into R as line feature data frame.

```{r}
cyclingpath <- st_read(dsn = "data/geospatial", 
                         layer = "CyclingPath")
```

The above message reveals that there are a total of 1625 Linestring features and 2 fields in 'CyclingPath' and it is in svy21 projected coordinates system too.

### 1.2.3 Importing GIS data in kml format

As the `pre-schools-location-kml` is in kml format, the syntax for importing the data differs slightly.

```{r}
preschool <- st_read("data/geospatial/pre-schools-location-kml.kml")
```
