---
title: "In Class Exericse 2"
author: "Hulwana"
editor: visual
execute:
  echo: false
  warning: false
  message: false
---

## Overview

## \## Getting Started

In the code chunk below, `p_load()` of *pacman* package is used to install and load the following R packages into R environment:

-   [sf](https://r-spatial.github.io/sf/) is use for importing and handling geospatial data in R,

-   [tidyverse](https://www.tidyverse.org/)is mainly use for wrangling attribute data in R,

-   [tmap](https://r-tmap.github.io/tmap/) will be used to prepare cartographic quality chropleth map,

-   [spdep](https://r-spatial.github.io/spdep/) will be used to compute spatial weights, global and local spatial autocorrelation statistics, and

-   [funModeling](https://cran.r-project.org/web/packages/funModeling/vignettes/funModeling_quickstart.html) will be used for rapid Exploratory Data Analysis

```{r}
pacman::p_load(sf, tidyverse, tmap, spdep, readr, tidyr)
```

## Importing Geospatial Data

In this in-class data, two geospatial datasets will beused, they are:

-   geo_export

-   nga_ADM2

### Importing Geospatial Data

First, we are going to import the water point geospatial data (i.e. geo_export) by using the code chunk below.

```{r}
#| eval: false
# wp <- st_read(dsn = "data", 
#                    layer = "geo_export",
#                    crs = 4326) %>%
#   filter(clean_coun == "Nigeria")

```

Things to learn from the code chunk above:

-   `st_read()` of **sf** package is used to import *geo_export* shapefile into R environment and save the imported geospatial data into simple feature data table.

-   `filter()` of dplyr package is used to extract water point records of Nigeria.

Next, `write_rds()` of readr package is used to save the extracted sf data table (i.e. wp) into an output file in rds data format. The output file is called *wp_nga.rds* and it is saved in *geodata* sub-folder.

```{r}
#write_rds(wp, "data/wp_nga.rds")
```

```{r}
wp_nga <- read_rds("data/wp_nga.rds") %>%
  dplyr::mutate(status_cle = 
           replace_na(status_cle, "Unknown"))
```

### Import Nigeria LGA Boundary data

Now, we are going to import the LGA boundary data into R environment by using the code chunk below.
