---
title: "In-class Exercise 1"
editor: visual
---

## Overview

For this In-class Exercise 1, we will import and prepare geospatial data using R.

## Getting Started

The code chunk below will install and load tidyverse and sf packages.

```{r}
pacman::p_load(sf, tidyverse)
```

## Importing Geospatial Data

### Importing polygon features

This code chunk will import ESRI shapefile into R. SVY21 is a Singapore projection system and the units is in metres. We also need to indicate the layer in the directory.

```{r}
mpsz <- st_read(dsn = "data/geospatial", layer = "MP14_SUBZONE_WEB_PL")
```

### Importing polyline feature data in shapefile form

The code chunk below uses *st_read()* function of **sf** package to import `CyclingPath` shapefile into R as line feature data frame. We need the projected coordinated system as we want to calculate the distance. The projected crs reflect the earth with different country having different coordinated systems.

```{r}
cyclingpath <- st_read(dsn = "data/geospatial", 
                         layer = "CyclingPath")
```

### Importing GIS data in kml format

As the `pre-schools-location-kml` is in kml format, the syntax for importing the data differs slightly.

```{r}
preschool <- st_read("data/geospatial/pre-schools-location-kml.kml")
```

As the above data is having a geodetic CRS of WGS 84 which is not efficient for geospatial analysis and therefore needs to be transformed into Singapore system of 3414.

### Importing and transforming into projected coordinates system

```{r}
preschool <- st_read("data/geospatial/pre-schools-location-kml.kml") %>%
  st_transform(crs = 3414)
```
